---
title: "Fig 2. Model comparisons"
author: "Sebastian Block et al."
output: html_document
---
## Fossil density in areas predicted by our models

This is the code needed to make figure 2 of our paper (Block *et al*., 2016), which shows how our models can identify areas where the probability of finding fossils is high.

The first step is to load the attribute table of the shapefile `3P.shp`. In the metadata for this file you can find an explanation of what each of the variables in the attribute table represent.

**NOTE**: Don't forget to set your working directory to the folder containing all the files, in this case, the folder `Block et al. (2015) Fossil Hunting`

```{r, results = 'hide', message=FALSE, warning=FALSE, eval=FALSE}
# The attribute table is a dbf file, and to load in R we require the "foreing" package
require(foreign)
gridFH <- read.dbf("../Data/Grids/3P.dbf")
```


Now, let's create a new dataframe that will contain the data needed to make the plot. 


```{r, eval=FALSE}
plot_data <- data.frame(Genus = rep(c("Diprotodon", "Zygomaturus", "Protemnodon", "Thylacoleo", "Genyornis"), each = 3), Area = rep(c("Preservation & Discovery", "Climate Envelope", "All combined"), 5), Proportion = numeric(15))

## Diprotodon
bkp <- nrow(subset(gridFH, DiproPAv == 1)) / nrow(gridFH) # Proportion of fossils in whole landscape

plot_data$Proportion[1] <- (nrow(subset(gridFH, DiproPAv ==1 & hotspots == 1)) / nrow(subset(gridFH, hotspots == 1))) / bkp # Proportion of fossils in FP & FD hotspots
plot_data$Proportion[2] <- (nrow(subset(gridFH, DiproPAv ==1 & dipropred == 1)) / nrow(subset(gridFH, dipropred == 1))) / bkp # Proportion of fossils in SDM
plot_data$Proportion[3] <- (nrow(subset(gridFH, DiproPAv ==1 & diprover == 1)) / nrow(subset(gridFH, diprover == 1))) / bkp

## Zygomaturus
bkp <- nrow(subset(gridFH, ZygoPAv == 1)) / nrow(gridFH) # Proportion of fossils in whole landscape

plot_data$Proportion[4] <- (nrow(subset(gridFH, ZygoPAv ==1 & hotspots == 1)) / nrow(subset(gridFH, hotspots == 1))) / bkp # Proportion of fossils in FP & FD hotspots
plot_data$Proportion[5] <- (nrow(subset(gridFH, ZygoPAv ==1 & zygopred == 1)) / nrow(subset(gridFH, zygopred == 1))) / bkp # Proportion of fossils in SDM
plot_data$Proportion[6] <- (nrow(subset(gridFH, ZygoPAv ==1 & zygover == 1)) / nrow(subset(gridFH, zygover == 1))) / bkp

## Protemnodon
bkp <- nrow(subset(gridFH, ProtePAv == 1)) / nrow(gridFH) # Proportion of fossils in whole landscape

plot_data$Proportion[7] <- (nrow(subset(gridFH, ProtePAv ==1 & hotspots == 1)) / nrow(subset(gridFH, hotspots == 1))) / bkp # Proportion of fossils in FP & FD hotspots
plot_data$Proportion[8] <- (nrow(subset(gridFH, ProtePAv ==1 & protepred == 1)) / nrow(subset(gridFH, protepred == 1))) / bkp # Proportion of fossils in SDM
plot_data$Proportion[9] <- (nrow(subset(gridFH, ProtePAv ==1 & protever == 1)) / nrow(subset(gridFH, protever == 1))) / bkp # Proportion of fossils in 3P

## Thylacoleo
bkp <- nrow(subset(gridFH, ThylaPAv == 1)) / nrow(gridFH) # Proportion of fossils in whole landscape

plot_data$Proportion[10] <- (nrow(subset(gridFH, ThylaPAv ==1 & hotspots == 1)) / nrow(subset(gridFH, hotspots == 1))) / bkp # Proportion of fossils in FP & FD hotspots
plot_data$Proportion[11] <- (nrow(subset(gridFH, ThylaPAv ==1 & thylapred == 1)) / nrow(subset(gridFH, thylapred == 1))) / bkp # Proportion of fossils in SDM
plot_data$Proportion[12] <- (nrow(subset(gridFH, ThylaPAv ==1 & thylaver == 1)) / nrow(subset(gridFH, thylaver == 1))) / bkp

## Genyornis
bkp <- nrow(subset(gridFH, GenyPAv == 1)) / nrow(gridFH) # Proportion of fossils in whole landscape

plot_data$Proportion[13] <- (nrow(subset(gridFH, GenyPAv ==1 & hotspots == 1)) / nrow(subset(gridFH, hotspots == 1))) / bkp # Proportion of fossils in FP & FD hotspots
plot_data$Proportion[14] <- (nrow(subset(gridFH, GenyPAv ==1 & genypred == 1)) / nrow(subset(gridFH, genypred == 1))) / bkp # Proportion of fossils in SDM
plot_data$Proportion[15] <- (nrow(subset(gridFH, GenyPAv ==1 & genyver == 1)) / nrow(subset(gridFH, genyver == 1))) / bkp

plot_data$Genus <- factor(plot_data$Genus, c("Diprotodon", "Zygomaturus", "Protemnodon", "Thylacoleo", "Genyornis"))
plot_data$Area <- factor(plot_data$Area, levels = c("Climate Envelope", "Preservation & Discovery", "All combined"))
```


The following code will create the figure and save it as a TIFF file in the specified directory. 


```{r, results = 'hide', message=FALSE, warning=FALSE, eval=FALSE}
# Load required packages and specify colors to use in plot
require(ggplot2)
require(grid)
cols <- c("#4cb0ff", "#7dbe84", "#aa0000") # This are the colors to use for the bars

# Make plot
tiff(filename = "../Display items/Fig 2 - Method comparison (v4).tiff",
    width = 15, height = 10, units = "cm", res = 300, compression = "lzw")
p <- ggplot(plot_data, aes(x=Genus, y=Proportion, fill=Area)) +
        geom_bar(stat="identity", position=position_dodge(), colour="black") +  
        scale_fill_manual(values = cols) + 
        theme(axis.title.x = element_blank(), text = element_text(size = 7.5)) +
        ylab("Probability of finding fossils (relative to random)") + expand_limits(y=c(0,10)) + theme_bw()
p + theme(legend.margin = unit(0, "cm"), axis.title.x = element_blank(), 
          axis.text.x  = element_text(face="italic"), legend.text = element_text(size = 7),
          panel.margin = unit(1, "cm"), panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(), legend.position = c(0.16, 0.85),
          legend.title = element_blank()) + geom_hline(aes(yintercept=1), linetype = "dashed")
dev.off()
```

```{r, echo = FALSE, fig.height = 3.94, fig.width = 4.72, eval=FALSE}
p <- ggplot(plot_data, aes(x=Genus, y=Proportion, fill=Area)) +
        geom_bar(stat="identity", position=position_dodge(), colour="black") +  
        scale_fill_manual(values = cols) + 
        theme(axis.title.x = element_blank(), text = element_text(size = 8)) +
        ylab("Fossil density compared to background") + expand_limits(y=c(0,15)) + theme_bw()
p + theme(legend.margin = unit(0, "cm"), axis.title.x = element_blank(), 
          axis.text.x  = element_text(face="italic"), legend.text = element_text(size = 8),
          panel.margin = unit(1, "cm"), panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(), legend.position = c(0.18, 0.86),
          legend.title = element_blank()) + geom_hline(aes(yintercept=1), linetype = "dashed")

```
