---
title: "Map Preparation"
author: "Sebastian Block et al."
date: "June 30, 2015"
output: html_document
---

This script is simply to prepare the model outputs to be used to create the final maps. <br><br>

```{r Loading and cleaning SDM, message = FALSE, eval=FALSE}
# Set working directory
setwd("~/Dropbox/Block et al 2015 Fossil Hunting") 

# Load attribute table of multiPCS.shp
require(foreign)
gridPCS <- read.dbf("Data/Grids/multiPCS.dbf")

# Rename columns with genus names (check that the genera are actually in this order)
names(gridPCS) <- c("Diprotodon", "Zygomaturus", "Protemnodon", "Thylacoleo", "Genyornis")
```

<br>
Since we lacked true absence data to calibate our models, model outputs can only be interpreted as a ranking of the suitability of potential of grid cells for species occurrence, rather than probabilities or likelihoods. Hence, we will transform the raw output into a real grid cell ranking, which will make the results more intuitive to interpret. <br> <br>

```{r Scaling probabilities SDM, eval=FALSE}
# Calculate ranks of SDM outputs
gridPCS$Dipro <- rank(gridPCS$Diprotodon) / nrow(gridPCS)
gridPCS$Zygo <- rank(gridPCS$Zygomaturus) / nrow(gridPCS)
gridPCS$Prote <- rank(gridPCS$Protemnodon) / nrow(gridPCS)
gridPCS$Thyla <- rank(gridPCS$Thylacoleo) / nrow(gridPCS)
gridPCS$Geny <- rank(gridPCS$Genyornis) / nrow(gridPCS)
```

<br>
Now, let's load and rank the outputs of the fossil preservation and discovery models. <br><br>

```{r Load & Scale GridProbs, eval=FALSE}
# Load attribute table of GridProbs.shp
gridPD <- read.dbf("Data/Grids/GridPD.dbf")
names(gridPD) <- c("Preservation", "Discovery")

# Rank outputs
gridPD$Preservation <- rank(gridPD$Preservation) / nrow(gridPD)
gridPD$Discovery <- rank(gridPD$Discovery) / nrow(gridPD)
```

<br>
Now, let's combine `gridPCS` and `gridPD`, calculate the final ranking of fossil potential (*i.e.*, the average of the rankings of palaeo-climate, preservation, and discovery suitabilities). <br><br>

```{r Final P, eval=FALSE}
# Combine data frames
gridPCS <- gridPCS[, 6:10]
FG <- cbind(gridPCS, gridPD)

# Calculate final (combined) fossil potential for each taxon
FG$diproP <- (FG$Dipro + FG$Preservation + FG$Discovery) / 3
FG$zygoP <- (FG$Zygo + FG$Preservation + FG$Discovery) / 3
FG$proteP <- (FG$Prote + FG$Preservation + FG$Discovery) / 3
FG$thylaP <- (FG$Thyla + FG$Preservation + FG$Discovery) / 3
FG$genyP <- (FG$Geny + FG$Preservation + FG$Discovery) / 3

# And rank them
FG$diproP <- rank(FG$diproP) / nrow(FG)
FG$zygoP <- rank(FG$zygoP) / nrow(FG)
FG$proteP <- rank(FG$proteP) / nrow(FG)
FG$thylaP <- rank(FG$thylaP) / nrow(FG)
FG$genyP <- rank(FG$genyP) / nrow(FG)
```

<br>
Finally, let's calculate the range of ranking predictions for different taxa in each grid cell, to see if making taxa-specific predictions makes sense or not. 
<br>

```{r Inter-taxa ranking range calculation, eval=FALSE}
# Palaeo-climate suitability
FG$rangePCS <- numeric(nrow(FG))

for (i in 1:nrow(FG)){
        highest <- max(FG$Dipro[i], FG$Zygo[i], FG$Prote[i], FG$Thyla[i], FG$Geny[i])
        lowest <- min(FG$Dipro[i], FG$Zygo[i], FG$Prote[i], FG$Thyla[i], FG$Geny[i])
        FG$rangePCS[i] <-  highest - lowest
        rm(highest, lowest)
}

summary(FG$rangePCS)

# Final fossil potential
FG$rangeFP <- numeric(nrow(FG))

for (i in 1:nrow(FG)){
        highest <- max(FG$diproP[i], FG$zygoP[i], FG$proteP[i], FG$thylaP[i], FG$genyP[i])
        lowest <- min(FG$diproP[i], FG$zygoP[i], FG$proteP[i], FG$thylaP[i], FG$genyP[i])
        FG$rangeFP[i] <-  highest - lowest
        rm(highest, lowest)
}
        
summary(FG$rangeFP)
```



```{r Write file, eval=FALSE}
# Write dbf file 
write.dbf(FG, "Data/Grids/multiPs.dbf")
```

