---
title: "Overlaps"
author: "Sebastian Block et al."
output: html_document
---

This script uses the continuous outputs predicted by the climate-envelope models (`multiPCS.shp`) and by the fossil preservation and discovery models (`GridPD.shp`) and converts them into a binary output with a **threshold that maximizes the sum of the sensitivity and specificity**. Then, we will calculate the overlap of those binary outputs (the areas with maximum potential of having fossils of a given taxon). <br><br>

## Data preparation


```{r Preparations, eval=FALSE}
# Load packages
require(foreign)
require(dismo)

# Set working directory
setwd("~/Dropbox/Block et al 2015 Fossil Hunting") 


taxaPA <- read.dbf("Data/Grids/taxaPA.dbf") # grid with density of taxa fossils in each grid cell
taxaPA$taxaPAv <- numeric(nrow(taxaPA))
taxaPA$taxaA <- numeric(nrow(taxaPA))

for (i in 1:nrow(taxaPA)) {
        taxaPA$taxaPAv[i] <- ifelse(max(taxaPA[i, 1:5]) > 0, 1, 0)
        taxaPA$taxaA[i] <- ifelse(max(taxaPA[i, 6:10]) > 0, 1, 0)
}

taxaPA[taxaPA$taxaA == 1, 11] <- 0 # All fossils

# Make taxa densities presence/absences
taxaPA$DiproPAv <- ifelse(taxaPA$DiproPAv > 0, 1, 0)
taxaPA$ZygoPAv <- ifelse(taxaPA$ZygoPAv > 0, 1, 0)
taxaPA$ProtePAv <- ifelse(taxaPA$ProtePAv > 0, 1, 0)
taxaPA$ThylaPAv <- ifelse(taxaPA$ThylaPAv > 0, 1, 0)
taxaPA$GenyPAv <- ifelse(taxaPA$GenyPAv > 0, 1, 0)

taxaPA$DiproA <- ifelse(taxaPA$DiproA > 0, 1, 0)
taxaPA$ZygoA <- ifelse(taxaPA$ZygoA > 0, 1, 0)
taxaPA$ProteA <- ifelse(taxaPA$ProteA > 0, 1, 0)
taxaPA$ThylaA <- ifelse(taxaPA$ThylaA > 0, 1, 0)
taxaPA$GenyA <- ifelse(taxaPA$GenyA > 0, 1, 0)

# Remove cells with both reliable and unreliable fossils of the testing
taxaPA[taxaPA$DiproA > 0, 1] <- 0 
taxaPA[taxaPA$ProteA > 0, 2] <- 0 
taxaPA[taxaPA$ThylaA > 0, 3] <- 0 
taxaPA[taxaPA$GenyoA > 0, 4] <- 0 
taxaPA[taxaPA$ZygoA > 0, 5] <- 0 

gridFP <- read.dbf("Data/Grids/Preservation.dbf")
taxaPA$fossPAtr <- ifelse(taxaPA$taxaPAv > 0, 0, gridFP$fossPA) # Exclude grid cells
rm(gridFP)


# Combine these presence / absence fields with data frame of model probabilities
gridPs <- read.dbf("Data/Grids/valPs.dbf")
gridPs <- cbind(gridPs, taxaPA)
```

<br>

## Fossil hotspots

Next, we will calculate thresholds for the predictions of the fossil preservation and fossil discovery models, and use the threshold to make binary predictions. Then, we will calculate the overlap of these two binary predictions. This will represent areas where fossils can be preserved *and* found. This are the fossil hotspots. <br><br>

```{r Fossil hotspots, eval=FALSE}
## Define presences and absences data
presences <- subset(gridPs, taxaPAv == 1)
pseudoabs <- subset(gridPs, taxaPAv == 0)

## Fossil preservation
mod_fp <- glm(fossPAtr ~ offset(Preservati), family="binomial", data = gridPs)
ep <- evaluate(presences, pseudoabs, mod_fp)
max(ep@TPR + ep@TNR) - 1
tp <- threshold(ep, 'spec_sens')
fppred <- predict(mod_fp)
fppred <- ifelse(fppred > tp, 1, 0)
gridPs$fppred <- fppred

## Fossil discovery
mod_fd <- glm(fossPAtr ~ offset(Discovery), family="binomial", data = gridPs)
ed <- evaluate(presences, pseudoabs, mod_fd) # Evaluate model (circular validation)
max(ed@TPR + ed@TNR) - 1 # Discovery model TSS
td <- threshold(ed, 'spec_sens') # Find threshold that maximizes TSS
fdpred <- predict(mod_fd) # Make model predictions
fdpred <- ifelse(fdpred > td, 1, 0) # Use threshold to make binary predictions
gridPs$fdpred <- fdpred

## Overlap of fossil discovery and preservation areas (fossil hotspots)
gridPs$hotspots <- gridPs$fppred * gridPs$fdpred
```

<br>

## Binary climate-envelope predictions

The fossil hotspots are the places to dig if one doesn't have preference for fossils of any particular species. If one does, it is better to calculate a binary output for the SDM of that species, and see where it overlaps whith the fossil hotspots. <br><br>


```{r SDMs areas, eval=FALSE}
# Diprotodon
presences <- subset(gridPs, DiproPAv == 1)
pseudoabs <- subset(gridPs, DiproPAv == 0)
pcem <- glm(DiproA ~ offset(Dipro), family="binomial", data = gridPs)
epcem <- evaluate(presences, pseudoabs, pcem)
max(epcem@TPR + epcem@TNR) - 1 # True Skill Statistic
tdip <- threshold(epcem, 'spec_sens')
dipropred <- predict(pcem)
dipropred <- ifelse(dipropred > tdip, 1, 0)
gridPs$dipropred <- dipropred
gridPs$diprover <- dipropred * gridPs$hotspots

# Zygomaturus
presences <- subset(gridPs, ZygoPAv == 1)
pseudoabs <- subset(gridPs, ZygoPAv == 0)
pcem <- glm(ZygoA ~ offset(Zygo), family="binomial", data = gridPs)
epcem <- evaluate(presences, pseudoabs, pcem)
max(epcem@TPR + epcem@TNR) - 1 # True Skill Statistic
tzyg <- threshold(epcem, 'spec_sens')
zygopred <- predict(pcem)
zygopred <- ifelse(zygopred > tzyg, 1, 0)
gridPs$zygopred <- zygopred
gridPs$zygover <- zygopred * gridPs$hotspots

# Protemnodon
presences <- subset(gridPs, ProtePAv == 1)
pseudoabs <- subset(gridPs, ProtePAv == 0)
pcem <- glm(ProteA ~ offset(Prote), family="binomial", data = gridPs)
epcem <- evaluate(presences, pseudoabs, pcem)
max(epcem@TPR + epcem@TNR) - 1 # True Skill Statistic
tprot <- threshold(epcem, 'spec_sens')
protepred <- predict(pcem)
protepred <- ifelse(protepred > tprot, 1, 0)
gridPs$protepred <- protepred
gridPs$protever <- protepred * gridPs$hotspots

# Thylacoleo
presences <- subset(gridPs, ThylaPAv == 1)
pseudoabs <- subset(gridPs, ThylaPAv == 0)
pcem <- glm(ThylaA ~ offset(Thyla), family="binomial", data = gridPs)
epcem <- evaluate(presences, pseudoabs, pcem)
max(epcem@TPR + epcem@TNR) - 1 # True Skill Statistic
tthy <- threshold(epcem, 'spec_sens')
thylapred <- predict(pcem)
thylapred <- ifelse(thylapred > tthy, 1, 0)
gridPs$thylapred <- thylapred
gridPs$thylaver <- thylapred * gridPs$hotspots

# Genyornis
presences <- subset(gridPs, GenyPAv == 1)
pseudoabs <- subset(gridPs, GenyPAv == 0)
pcem <- glm(GenyA ~ offset(Geny), family="binomial", data = gridPs)
epcem <- evaluate(presences, pseudoabs, pcem)
max(epcem@TPR + epcem@TNR) - 1 # True Skill Statistic
tgen <- threshold(epcem, 'spec_sens')
genypred <- predict(pcem)
genypred <- ifelse(genypred > tgen, 1, 0)
gridPs$genypred <- genypred
gridPs$genyver <- genypred * gridPs$hotspots
```

<br>
Finally, let's save all our calculations in a new shapefile called `3P.shp`

```{r Saving dbf, eval=FALSE}
write.dbf(gridPs, "Data/Grids/3P.dbf")
```

