---
title: "Validations"
author: "Sebastian Block et al."
output: html_document
---

## Climate-envelope model validations

Let's start by comparing the predictions of the climate-envelope models in grid cells with non-relliably dated fossils (not used for model training) and in random grid cells. First we will compare predicted suitability values in grid cells with fossils versus values in grid cells selected randomly with the Kolmogorov-Smirnov test. <br><br>

```{r KS Validation SDM, eval=FALSE}
# Set working directory
setwd("~/Dropbox/Block et al 2015 Fossil Hunting") 

# Load packages
require(foreign)
require(ggplot2)

# Load data
gridPs <- read.dbf("Data/Grids/valPs.dbf")
taxaPA <- read.dbf("Data/Grids/taxaPA.dbf") # grid with density of taxa fossils in each grid cell

# Convert densities into presence (1) / absences (0)
## Validation fossils (unreliable ages)
taxaPA$DiproPAv <- ifelse(taxaPA$DiproPAv > 0, 1, 0)
taxaPA$ZygoPAv <- ifelse(taxaPA$ZygoPAv > 0, 1, 0)
taxaPA$ProtePAv <- ifelse(taxaPA$ProtePAv > 0, 1, 0)
taxaPA$ThylaPAv <- ifelse(taxaPA$ThylaPAv > 0, 1, 0)
taxaPA$GenyPAv <- ifelse(taxaPA$GenyPAv > 0, 1, 0)


## Training fossils (reliable ages)
taxaPA$DiproA <- ifelse(taxaPA$DiproA > 0, 1, 0)
taxaPA$ZygoA <- ifelse(taxaPA$ZygoA > 0, 1, 0)
taxaPA$ProteA <- ifelse(taxaPA$ProteA > 0, 1, 0)
taxaPA$ThylaA <- ifelse(taxaPA$ThylaA > 0, 1, 0)
taxaPA$GenyA <- ifelse(taxaPA$GenyA > 0, 1, 0)

gridPs <- cbind(gridPs, taxaPA)

# Since the validation includes random selection of grid cells, to reproduce the exact same results
# reported in the paper, it's necessary to set a seed.
set.seed(8)

## Diprotodon
# Values of validation fossils
pred.foss <- subset(gridPs, DiproPAv == 1 & DiproA == 0)$Dipro
background <- subset(gridPs, DiproPAv == 0 & DiproA == 0)$Dipro

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)

####

## Zygomaturus
pred.foss <- subset(gridPs, ZygoPAv == 1 & ZygoA == 0)$Zygo
background <- subset(gridPs, ZygoPAv == 0 & ZygoA == 0)$Zygo

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)

####

## Protemnodon
# Values of validation fossils
pred.foss <- subset(gridPs, ProtePAv == 1 & ProteA == 0)$Prote
background <- subset(gridPs, ProtePAv == 0 & ProteA == 0)$Prote

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)
####

## Thylacoleo
# Values of validation fossils
pred.foss <- subset(gridPs, ThylaPAv == 1 & ThylaA == 0)$Thyla
background <- subset(gridPs, ThylaPAv == 0 & ThylaA == 0)$Thyla

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)
####

## Genyornis
pred.foss <- subset(gridPs, GenyPAv == 1 & GenyA == 0)$Geny
background <- subset(gridPs, GenyPAv == 0 & GenyA == 0)$Geny

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)
```
<br>

Now let's calculate the true skill statistic.

<br>

```{r TSS Validation SDMs, eval=FALSE}
set.seed(8)
require(dismo)
# Diprotodon
tss <- numeric(1000)
for (i in 1:1000){
        presences <- gridPs[gridPs$DiproPAv == 1 & gridPs$DiproA == 0, "Dipro"]
        pseudoabs <- gridPs[gridPs$DiproPAv == 0 & gridPs$DiproA == 0, "Dipro"]
        pseudoabs <- pseudoabs[sample(length(pseudoabs), length(presences)*10)]
        metasdm <- glm(DiproPAv ~ offset(Dipro) - 1, family="binomial", data = gridPs)
        eva <- evaluate(presences, pseudoabs, metasdm)
        tss[i] <- max(eva@TPR + eva@TNR) - 1
}
summary(tss)

# Zygomaturus
tss <- numeric(1000)
for (i in 1:1000){
        presences <- gridPs[gridPs$GenyPAv == 1 & gridPs$ZygoA == 0, "Zygo"]
        pseudoabs <- gridPs[gridPs$GenyPAv == 0 & gridPs$ZygoA == 0, "Zygo"]
        pseudoabs <- pseudoabs[sample(length(pseudoabs), length(presences)*10)]
        metasdm <- glm(GenyPAv ~ offset(Geny) - 1, family="binomial", data = gridPs)
        eva <- evaluate(presences, pseudoabs, metasdm)
        tss[i] <- max(eva@TPR + eva@TNR) - 1
}
summary(tss)


# Protemnodon
tss <- numeric(1000)
for (i in 1:1000){
        presences <- gridPs[gridPs$ProtePAv == 1 & gridPs$ProteA == 0, "Prote"]
        pseudoabs <- gridPs[gridPs$ProtePAv == 0 & gridPs$ProteA == 0, "Prote"]
        pseudoabs <- pseudoabs[sample(length(pseudoabs), length(presences)*10)]
        metasdm <- glm(ProtePAv ~ offset(Prote) - 1, family="binomial", data = gridPs)
        eva <- evaluate(presences, pseudoabs, metasdm)
        tss[i] <- max(eva@TPR + eva@TNR) - 1
}
summary(tss)

# Thylacoleo
tss <- numeric(1000)
for (i in 1:1000){
        presences <- gridPs[gridPs$ThylaPAv == 1 & gridPs$ThylaA == 0, "Thyla"]
        pseudoabs <- gridPs[gridPs$ThylaPAv == 0 & gridPs$ThylaA == 0, "Thyla"]
        pseudoabs <- pseudoabs[sample(length(pseudoabs), length(presences)*10)]
        metasdm <- glm(ThylaPAv ~ offset(Thyla) - 1, family="binomial", data = gridPs)
        eva <- evaluate(presences, pseudoabs, metasdm)
        tss[i] <- max(eva@TPR + eva@TNR) - 1
}
summary(tss)

# Genyornis
tss <- numeric(1000)
for (i in 1:1000){
        presences <- gridPs[gridPs$GenyPAv == 1 & gridPs$GenyA == 0, "Geny"]
        pseudoabs <- gridPs[gridPs$GenyPAv == 0 & gridPs$GenyA == 0, "Geny"]
        pseudoabs <- pseudoabs[sample(length(pseudoabs), length(presences)*10)]
        metasdm <- glm(GenyPAv ~ offset(Geny) - 1, family="binomial", data = gridPs)
        eva <- evaluate(presences, pseudoabs, metasdm)
        tss[i] <- max(eva@TPR + eva@TNR) - 1
}
summary(tss)
```

<br>

Now let's do the same for the preservation model. 


```{r KS Validation Preservation, eval=FALSE}
set.seed(8)

## Diprotodon
# Values of validation fossils
pred.foss <- subset(gridPs, DiproPAv == 1 & DiproA == 0)$Preservati
background <- subset(gridPs, DiproPAv == 0 & DiproA == 0)$Preservati

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)

####

## Zygomaturus
pred.foss <- subset(gridPs, ZygoPAv == 1 & ZygoA == 0)$Preservati
background <- subset(gridPs, ZygoPAv == 0 & ZygoA == 0)$Preservati

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)

####

## Protemnodon
# Values of validation fossils
pred.foss <- subset(gridPs, ProtePAv == 1 & ProteA == 0)$Preservati
background <- subset(gridPs, ProtePAv == 0 & ProteA == 0)$Preservati

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)
####

## Thylacoleo
# Values of validation fossils
pred.foss <- subset(gridPs, ThylaPAv == 1 & ThylaA == 0)$Preservati
background <- subset(gridPs, ThylaPAv == 0 & ThylaA == 0)$Preservati

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)
####

## Genyornis
pred.foss <- subset(gridPs, GenyPAv == 1 & GenyA == 0)$Preservati
background <- subset(gridPs, GenyPAv == 0 & GenyA == 0)$Preservati

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)
```

And discovery model.

```{r KS Validation Discovery, eval=FALSE}
set.seed(8)

## Diprotodon
# Values of validation fossils
pred.foss <- subset(gridPs, DiproPAv == 1 & DiproA == 0)$Discovery
background <- subset(gridPs, DiproPAv == 0 & DiproA == 0)$Discovery

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)

####

## Zygomaturus
pred.foss <- subset(gridPs, ZygoPAv == 1 & ZygoA == 0)$Discovery
background <- subset(gridPs, ZygoPAv == 0 & ZygoA == 0)$Discovery

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)

####

## Protemnodon
# Values of validation fossils
pred.foss <- subset(gridPs, ProtePAv == 1 & ProteA == 0)$Discovery
background <- subset(gridPs, ProtePAv == 0 & ProteA == 0)$Discovery

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)
####

## Thylacoleo
# Values of validation fossils
pred.foss <- subset(gridPs, ThylaPAv == 1 & ThylaA == 0)$Discovery
background <- subset(gridPs, ThylaPAv == 0 & ThylaA == 0)$Discovery

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)
####

## Genyornis
pred.foss <- subset(gridPs, GenyPAv == 1 & GenyA == 0)$Discovery
background <- subset(gridPs, GenyPAv == 0 & GenyA == 0)$Discovery

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)
```

And the overlap of preservation and discovery models.

```{r KS Validation Fossil Hotspots, eval=FALSE}
set.seed(8)

## Diprotodon
# Values of validation fossils
pred.foss1 <- subset(gridPs, DiproPAv == 1 & DiproA == 0)$Discovery
pred.foss2 <- subset(gridPs, DiproPAv == 1 & DiproA == 0)$Preservati
pred.foss <- (pred.foss1 + pred.foss2) / 2
background1 <- subset(gridPs, DiproPAv == 0 & DiproA == 0)$Discovery
background2 <- subset(gridPs, DiproPAv == 0 & DiproA == 0)$Preservati
background <- (background1 + background2) / 2

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)

####

## Zygomaturus
pred.foss1 <- subset(gridPs, ZygoPAv == 1 & ZygoA == 0)$Discovery
pred.foss2 <- subset(gridPs, ZygoPAv == 1 & ZygoA == 0)$Preservati
pred.foss <- (pred.foss1 + pred.foss2) / 2
background1 <- subset(gridPs, ZygoPAv == 0 & ZygoA == 0)$Discovery
background2 <- subset(gridPs, ZygoPAv == 0 & ZygoA == 0)$Preservati
background <- (background1 + background2) / 2

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)

####

## Protemnodon
# Values of validation fossils
pred.foss1 <- subset(gridPs, ProtePAv == 1 & ProteA == 0)$Discovery
pred.foss2 <- subset(gridPs, ProtePAv == 1 & ProteA == 0)$Preservati
pred.foss <- (pred.foss1 + pred.foss2) / 2
background1 <- subset(gridPs, ProtePAv == 0 & ProteA == 0)$Discovery
background2 <- subset(gridPs, ProtePAv == 0 & ProteA == 0)$Preservati
background <- (background1 + background2) / 2

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)
####

## Thylacoleo
# Values of validation fossils
pred.foss1 <- subset(gridPs, ThylaPAv == 1 & ThylaA == 0)$Discovery
pred.foss2 <- subset(gridPs, ThylaPAv == 1 & ThylaA == 0)$Preservati
pred.foss <- (pred.foss1 + pred.foss2) / 2
background1 <- subset(gridPs, ThylaPAv == 0 & ThylaA == 0)$Discovery
background2 <- subset(gridPs, ThylaPAv == 0 & ThylaA == 0)$Preservati
background <- (background1 + background2) / 2

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)
####

## Genyornis
pred.foss1 <- subset(gridPs, GenyPAv == 1 & GenyA == 0)$Discovery
pred.foss2 <- subset(gridPs, GenyPAv == 1 & GenyA == 0)$Preservati
pred.foss <- (pred.foss1 + pred.foss2) / 2
background1 <- subset(gridPs, GenyPAv == 0 & GenyA == 0)$Discovery
background2 <- subset(gridPs, GenyPAv == 0 & GenyA == 0)$Preservati
background <- (background1 + background2) / 2

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)

```


And the overlap of the three models (climate-envelope, preservation, and discovery)

```{r KS Validation Final P, eval=FALSE}
set.seed(8)

## Diprotodon
# Values of validation fossils
pred.foss <- subset(gridPs, DiproPAv == 1 & DiproA == 0)$diproP
background <- subset(gridPs, DiproPAv == 0 & DiproA == 0)$diproP

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)

####

## Zygomaturus
pred.foss <- subset(gridPs, ZygoPAv == 1 & ZygoA == 0)$zygoP
background <- subset(gridPs, ZygoPAv == 0 & ZygoA == 0)$zygoP

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)

####

## Protemnodon
# Values of validation fossils
pred.foss <- subset(gridPs, ProtePAv == 1 & ProteA == 0)$proteP
background <- subset(gridPs, ProtePAv == 0 & ProteA == 0)$proteP

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)
####

## Thylacoleo
# Values of validation fossils
pred.foss <- subset(gridPs, ThylaPAv == 1 & ThylaA == 0)$thylaP
background <- subset(gridPs, ThylaPAv == 0 & ThylaA == 0)$thylaP

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)
####

## Genyornis
pred.foss <- subset(gridPs, GenyPAv == 1 & GenyA == 0)$genyP
background <- subset(gridPs, GenyPAv == 0 & GenyA == 0)$genyP

# Compare to 
kstats <- numeric(1000)
for(i in 1:1000){
        random_pts <- sample(background, length(pred.foss))
        kstats[i] <- ks.test(pred.foss, random_pts, alternative = "less")$statistic
}
summary(kstats)
```

